Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2018-01-15 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
change dominance of restoration to dominance of resistance

Parameter  | Description
---------------------- | ----------------------------------------------------
1. Effectiveness | proportion of susceptible (SS) insects killed by exposure to insecticide
2. Exposure | proportion of insects exposed to insecticide
3. Resistance restoration  | ability of resistance (RR) to restore fitness of insects exposed to insecticide
4. Dominance of restoration | sets fitness of heterozygous (SR) insects between that of SS & RR in presence of insecticide
5. Frequency  | frequency of resistance alleles within the population
6. Cost of resistance  |  fitness of resistant (RR) insects in absence of insecticide
7. Dominance of cost  | sets fitness of heterozygous (SR) insects between that of SS & RR in absence of insecticide

@@

Dominance of restoration determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Fig 2).

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and dominance of restoration all resulted in faster resistance spread (Fig 4 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 5A) through the even simpler mechanism that a smaller change in frequency is required to reach the threshold.

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ------ | ---------------------------------------------
1. Effectiveness | faster | reduced fitness of SS & SR exposed to insecticide
2. Exposure | faster | reduced fitness of SS & SR overall
3. Dominance of restoration | faster | increased fitness of SR exposed to insecticide
4. Resistance restoration | faster | increased fitness of RR & SR exposed to insecticide
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of RR & SR not exposed to insecticide

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ---------- | ---------------------------------------------
1. Effectiveness | **slower** | one insecticide reduces the fitness of individuals resistant to the other thereby reducing selection pressure for the other
2. Exposure | faster (but less than for insecticide used alone) | reduced fitness of individuals susceptible to one insecticide increases selection pressure for that resistance. However at the same time selection pressure is reduced by reduced fitness of resistant individuals caused by the other insecticide
3. Dominance of restoration | faster | increased fitness of heterozygotes
4. Resistance restoration | faster | increased fitness of resistants
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of resistants in absence of insecticide

@@

Parameter to increase  | increase favours mix or sequence | Mechanism
----------------------- | -------- | ---------------------------------------------
1. Effectiveness | mixture | Higher effectiveness gives faster resistance for sequence and slower resistance in mixture
2. Exposure | sequence | Higher exposure gives faster resistance for sequence and mixture but the greater effect on mixture favours sequence.
3. Dominance of restoration | neither | Higher dominance gives faster resistance in both sequences and mixtures such that the difference between them is not changed. 
4. Resistance restoration | neither | As for dominance of restoration. 
5. Frequency  | neither | As for dominance and resistance restoration. 
6. Cost of resistance | mixture | Higher costs slow the evolution of resistance more in a mixture than when used in sequence (but with higher costs in a sequence there is a greater chance for resistance levels to decline for the insecticide not being used).

@@

Parameter  | Calculation 
---------------------- | ----------------------------------------------------
1. Effectiveness | 1 - SSfit in presence of insecticide
2. Exposure | estimated
3. Resistance restoration  | (RRfit-SSfit) / Effectiveness in presence of insecticide
4. Dominance of restoration | (SRfit-SSfit)/(RRfit-SSfit) in presence of insecticide
5. Frequency  | estimated
6. Cost of resistance  |  1 - RRfit in absence of insecticide
7. Dominance of cost  | (SRfit-RRfit)/(RRfit-SSfit) in absence of insecticide

@@

**Field data**|Pyrethroid / Kdr[@Kolaczinski2000]|rescaled Pyrethroid / Kdr[@Kolaczinski2000]|Carbamate /  Ace1[@Essandoh2013]
------------------------------|-------|-------|-------
exposed survival RR|0.83|1|0.84
exposed survival SR|0.72|0.87|0.56
exposed survival SS|0.59|0.71|0.02
unexposed survival RR|0.67|0.81||-
unexposed survival SR|0.50|0.60|-
unexposed survival SS|0.83|1|-
**Calculated model inputs**|    |    |
Effectiveness|0.41|0.29|0.98
Resistance restoration|0.60|1|0.84
Dominance of restoration|0.52|0.54|0.66
Cost of resistance|-|0.19|-
Dominance of cost|-|1*|-

@@

When single insecticide use was investigated (Figs 4-5) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution of resistance, whereas higher values for cost led to slower evolution of resistance (Fig 5B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to Figure 2 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

@@

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? There are two main potential explanations for this apparent difference, a) dominance and b) polygenic resistance. The effect of dominance is that poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 4C, 8). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research on how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer. Secondly, if resistance is predominantly polygenic rather than monogenic poor implementation is likely to promote resistance by allowing the build up of many genes of small effect [@Gressel2017]. Indeed modelling work from 15 years ago [@Gardner1998][@Gardner1999] proposes an alternation of low and high doses with the low doses limiting the evolution of monogenic resistance and the high doses limiting that of the polygenic resistance. More research on the relative importance of monogenic and polygenic resistance for vector control would help here.

@@

```{r Fig2, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Figure 2. The effect of model inputs on the fitness of genotypes for a single insecticide. Fitness is shown on the y-axis and the different genotypes (SS, SR, RR) on the x axis. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS, resistance restoration 'restores' a portion of the fitness for RR and dominance of restoration determines how the fitness for SR lies between that of SS and RR. For those that are not exposed, fitness of SS is set to 1 by definition, resistance cost determines the fitness of RR and again dominance of cost determines how the fitness for SR sits between that of SS and RR. In this example effectiveness=0.8, resistance restoration=0.5 which 'restores' half of the fitness lost due to the insecticide, dominance of restoration=0.7 which sets the fitness of the SR closer to RR than SS. Resistance cost=0.3 which reduces fitness in the absence of the insecticide from 1 to 0.7, and dominance of cost=0.8 which sets fitness of SR close to RR."}

@@

```{r Fig4, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7.5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Figure 4. Single insecticide effects of effectiveness, exposure, dominance and resistance restoration on resistance frequency. A. Effectiveness, B. Exposure, C. Dominance of restoration, D. Resistance restoration. Increasing any of the inputs (going from red to blue) results in shorter times-to-resistance. For each panel the chosen input was varied in isolation with the remaining inputs set to 0.5, except for starting frequency of resistance which was set to 0.01). "}

@@

  #C 
  plotlist[[3]] <-ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(h.RS1_A0))) + 
         theme_bw() +
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) + 
         labs(colour = "dominance of restoration") +
         coord_trans(y = "log10") + 
         geom_line()

>>>>

Parameter  | Description
---------------------- | ----------------------------------------------------
1. Effectiveness | proportion of susceptible (SS) insects killed by exposure to insecticide
2. Exposure | proportion of insects exposed to insecticide
3. Resistance restoration  | ability of resistance (RR) to restore fitness of insects exposed to insecticide
4. Dominance of resistance | sets fitness of heterozygous (SR) insects between that of SS & RR in presence of insecticide
5. Frequency  | frequency of resistance alleles within the population
6. Cost of resistance  |  fitness of resistant (RR) insects in absence of insecticide
7. Dominance of cost  | sets fitness of heterozygous (SR) insects between that of SS & RR in absence of insecticide

@@

Dominance of resistance determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Fig 2).

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and dominance of resistance all resulted in faster resistance spread (Fig 4 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 5A) through the even simpler mechanism that a smaller change in frequency is required to reach the threshold.

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ------ | ---------------------------------------------
1. Effectiveness | faster | reduced fitness of SS & SR exposed to insecticide
2. Exposure | faster | reduced fitness of SS & SR overall
3. Dominance of resistance | faster | increased fitness of SR exposed to insecticide
4. Resistance restoration | faster | increased fitness of RR & SR exposed to insecticide
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of RR & SR not exposed to insecticide

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ---------- | ---------------------------------------------
1. Effectiveness | **slower** | one insecticide reduces the fitness of individuals resistant to the other thereby reducing selection pressure for the other
2. Exposure | faster (but less than for insecticide used alone) | reduced fitness of individuals susceptible to one insecticide increases selection pressure for that resistance. However at the same time selection pressure is reduced by reduced fitness of resistant individuals caused by the other insecticide
3. Dominance of resistance | faster | increased fitness of heterozygotes
4. Resistance restoration | faster | increased fitness of resistants
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of resistants in absence of insecticide

@@

Parameter to increase  | increase favours mix or sequence | Mechanism
----------------------- | -------- | ---------------------------------------------
1. Effectiveness | mixture | Higher effectiveness gives faster resistance for sequence and slower resistance in mixture
2. Exposure | sequence | Higher exposure gives faster resistance for sequence and mixture but the greater effect on mixture favours sequence.
3. Dominance of resistance | neither | Higher dominance gives faster resistance in both sequences and mixtures such that the difference between them is not changed. 
4. Resistance restoration | neither | As for dominance of resistance. 
5. Frequency  | neither | As for dominance and resistance restoration. 
6. Cost of resistance | mixture | Higher costs slow the evolution of resistance more in a mixture than when used in sequence (but with higher costs in a sequence there is a greater chance for resistance levels to decline for the insecticide not being used).

@@

Parameter  | Calculation 
---------------------- | ----------------------------------------------------
1. Effectiveness | 1 - SSfit in presence of insecticide
2. Exposure | estimated
3. Resistance restoration  | (RRfit-SSfit) / Effectiveness in presence of insecticide
4. Dominance of resistance | (SRfit-SSfit)/(RRfit-SSfit) in presence of insecticide
5. Frequency  | estimated
6. Cost of resistance  |  1 - RRfit in absence of insecticide
7. Dominance of cost  | (SRfit-RRfit)/(RRfit-SSfit) in absence of insecticide

@@

**Field data**|Pyrethroid / Kdr[@Kolaczinski2000]|rescaled Pyrethroid / Kdr[@Kolaczinski2000]|Carbamate /  Ace1[@Essandoh2013]
------------------------------|-------|-------|-------
exposed survival RR|0.83|1|0.84
exposed survival SR|0.72|0.87|0.56
exposed survival SS|0.59|0.71|0.02
unexposed survival RR|0.67|0.81||-
unexposed survival SR|0.50|0.60|-
unexposed survival SS|0.83|1|-
**Calculated model inputs**|    |    |
Effectiveness|0.41|0.29|0.98
Resistance restoration|0.60|1|0.84
Dominance of resistance|0.52|0.54|0.66
Cost of resistance|-|0.19|-
Dominance of cost|-|1*|-

@@

When single insecticide use was investigated (Figs 4-5) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution of resistance, whereas higher values for cost led to slower evolution of resistance (Fig 5B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to Figure 2 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of resistance acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

@@

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? There are two main potential explanations for this apparent difference, a) dominance and b) polygenic resistance. The effect of dominance is that poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of resistance leads to faster spread of resistance (Fig 4C, 8). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research on how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer. Secondly, if resistance is predominantly polygenic rather than monogenic poor implementation is likely to promote resistance by allowing the build up of many genes of small effect [@Gressel2017]. Indeed modelling work from 15 years ago [@Gardner1998][@Gardner1999] proposes an alternation of low and high doses with the low doses limiting the evolution of monogenic resistance and the high doses limiting that of the polygenic resistance. More research on the relative importance of monogenic and polygenic resistance for vector control would help here.

@@

```{r Fig2, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Figure 2. The effect of model inputs on the fitness of genotypes for a single insecticide. Fitness is shown on the y-axis and the different genotypes (SS, SR, RR) on the x axis. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS, resistance restoration 'restores' a portion of the fitness for RR and dominance of resistance determines how the fitness for SR lies between that of SS and RR. For those that are not exposed, fitness of SS is set to 1 by definition, resistance cost determines the fitness of RR and again dominance of cost determines how the fitness for SR sits between that of SS and RR. In this example effectiveness=0.8, resistance restoration=0.5 which 'restores' half of the fitness lost due to the insecticide, dominance of resistance=0.7 which sets the fitness of the SR closer to RR than SS. Resistance cost=0.3 which reduces fitness in the absence of the insecticide from 1 to 0.7, and dominance of cost=0.8 which sets fitness of SR close to RR."}

@@

```{r Fig4, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7.5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Figure 4. Single insecticide effects of effectiveness, exposure, dominance and resistance restoration on resistance frequency. A. Effectiveness, B. Exposure, C. dominance of resistance, D. Resistance restoration. Increasing any of the inputs (going from red to blue) results in shorter times-to-resistance. For each panel the chosen input was varied in isolation with the remaining inputs set to 0.5, except for starting frequency of resistance which was set to 0.01). "}

@@

  #C 
  plotlist[[3]] <-ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(h.RS1_A0))) + 
         theme_bw() +
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) + 
         labs(colour = "dominance of resistance") +
         coord_trans(y = "log10") + 
         geom_line()
